---
title: "Exercise 6"
output: html_document
date: "2025-02-24"
---

For this exercise, you will generate graphs that show univariate, descriptive statistics for nominal, ordinal, and continuous variables.
```{r}
#Use this code chunk to download the NHANES 2017-2018 database and to label the nominal and ordinal variables as factor variables (if necessary):
NHANESdata <- read.csv("https://raw.githubusercontent.com/malenamalka/ANTH241-SP25/refs/heads/main/data/NHANES2017-28")
factor(c(NHANESdata$Gender, NHANESdata$Race, NHANESdata$HealthIns))
factor(c(NHANESdata$Education,NHANESdata$HHEducation,NHANESdata$HHIncome,NHANESdata$FeelBad,NHANESdata$HHFoodSec),order=TRUE)
factor(NHANESdata$BMIcat, order =TRUE, levels = c("Underweight","Healthy","Overweight","Obese"))
head(NHANESdata)
```

1. Load the following packages to complete this exercise: ggplot2, tidyr, and dplyr.
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
```

1. Bar charts illustrate variation for nominal and ordinal variables.
```{r}
#a. Bar chart for a nominal variable showing raw counts: Create the bar chart for BMI categories (BMIcat)
#ggplot (data, aes(x=Catvar)) + geom_bar() + labs(title="", x="", y="")

ggplot(NHANESdata, aes(x=BMIcat)) + geom_bar(fill = "pink") +labs(title="BMI categories", x="BMIcat", y="counts")

```

```{r}
#b. Bar chart showing BMI categories by percentages:
#new_table_name <- with(NHANESdata, table(BMIcat))
BMIcat_table<- with (NHANESdata, table (BMIcat))
#turn the above table into proportions: prop.table(new_table_name)
prop.table(BMIcat_table)
#turn the table into percentages: prop.table(new_table_name)*100
prop.table(BMIcat_table)*100
#new_data_frame <- as.data.frame(prop.table(new_table_name)*100
BMIcat_barchart <- as.data.frame(prop.table(BMIcat_table)*100)
#ggplot(new_data_frame, aes(BMIcat, Freq)) + geom_col()
ggplot(BMIcat_barchart, aes(BMIcat, Freq)) + geom_col()


```

c. Construct bar charts for one ordinal/nominal variable in the code chunk below:
```{r}
#hheducaiton, hhincome, example a

ggplot(NHANESdata, aes(x=HHEducation))+geom_bar()+labs(title="Household education bar chart Bar Chart", x="HHEducation", y="Counts")

```

d. Constructing bar charts for a nominal/ordinal variable grouped by another nominal/ordinal variable:
```{r}
#BMI_cats by gender: stacked, grouped and filled bar charts:
#table(data$Independent variable, data$Dependent variable)

table(NHANESdata$Gender, NHANESdata$BMIcat)

#Stacked bar chart:
#ggplot(data, aes(x=Independent variable, fill=Dependent variable)) + geom_bar(position = "stack") + labs(title="", x="", y="")

ggplot(NHANESdata, aes(x=BMIcat, fill=Gender)) + geom_bar(position = "stack") + labs(title="stacked bar chart", x="BMIcat by gender", y="counts")

ggplot(NHANESdata, aes(x=BMIcat, fill=Gender)) + geom_bar(position = "stack") + facet_wrap(~Race)+ labs(title="stacked bar chart", x="BMIcat by gender", y="counts")

ggplot(NHANESdata, aes(x=BMIcat, fill=Gender)) + geom_bar(position = "stack") + facet_grid(~Race)+ labs(title="stacked bar chart", x="BMIcat by gender", y="counts")

#grouped bar chart:
#ggplot(data, aes(x=Independent variable, fill=dependent variable)) + geom_bar(position = "dodge") + labs(title="", x="Gender", y="Frequency")

ggplot(NHANESdata, aes(x=BMIcat, fill=Gender)) +geom_bar(position= "dodge") + labs(tite="Grouped bar chart", x="Gender", y="Frequency")

ggplot(NHANESdata, aes(x=BMIcat, fill=Gender)) +geom_bar(position= "dodge") + facet_wrap(~Race)+ labs(tite="Grouped bar chart", x="Gender", y="Frequency")

ggplot(NHANESdata, aes(x=BMIcat, fill=Gender)) +geom_bar(position= "dodge") + facet_grid(~Race)+ labs(tite="Grouped bar chart", x="Gender", y="Frequency")

```

e. Constructing bar charts for mean +/- standard error of continuous variables
```{r}
#1 generate means and standard deviation of BMI(dependent var.) by gender (independent var)
#new_data_frame <- data%>%
#drop_na(Dependent variable)%>%
#group_by(Independent variable)
#summarize(n=n(), mean=mean(dependent variable), sd=sd(dependent variable))
#to print new data frame: new_data_frame

BMI_gender <- NHANESdata%>%
  drop_na(BMI)%>%
  group_by(Gender)%>%
  summarize(n=n(), mean=mean(BMI), sd=sd(BMI))
BMI_gender

#2 Plot means in a sorted bar chart and adds error bars that represent the standard deviation
#ggplot(new_data_frame, aes(x=reorder(Independent variable, mean), y=mean)) + geom_bar(stat="identity", fill="color") + geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.2)
#labs(x="", y="", title="", subtitle="")

ggplot(BMI_gender, aes(x=reorder(Gender, mean), y=mean))+ geom_bar(stat="identity", fill="pink") +geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.2) + labs(x="mean BMI by gender", title="Mean BMI", subtitle="with standard deviation")
```

2. Histograms illustrate variation in continuous variables:
```{r}

ggplot(NHANESdata, aes(x=BMI)) +
  geom_histogram() +labs(title="Histogram of BMI")

```

```{r}
#b. colored histogram with 20 bins: ggplot(data, aes(x=contvar)) + geom_histogram(bins=20, color="white", fill="steelblue") + labs(title="Colored histogram with 20 bins", x="", y="")
#do the above code for BMI:

ggplot(NHANESdata, aes(x=BMI)) +
  geom_histogram(bins=50, color="skyblue", fill="pink")+
  labs(title="colored histogram with 50 bins", x="BMI", y="Count")

```

```{r}
#c. Histogram with percentages:
#ggplot(data, aes(x=contvar, y=..density..)) + geom_histogram(bins=20, color="white", fill="steelblue") + scale_y_continuous(labels=scales::percent) + labs(title="", y="percent", x="")

ggplot(NHANESdata, aes(x=BMI, y=..density..))+geom_histogram(bins=20, color="blue", fill="pink")+ scale_y_continuous(labels=scales::percent)+labs(title="histogram with percentages", y="percent", x="BMI")

```

d: Create a histogram of a different continuous variable:

```{r}

ggplot(NHANESdata, aes(DR1TKCAL))+geom_histogram() +labs(title="Calorie intake histogram", x="Calories", y="Count")
```


3. Box plots: a box and whisker plot describes the distribution of a continuous variable by plotting its five-number summary: minimum, lower quartile (25th percentile), median (50th percentile), upper quartile (75th percentile), and maximum, and the outliers (values outside of the +/- 1.5xQR)
```{r}
#a.ggplot(data, aes(x="", y=Contvar)) + geom_boxplot() + labs(y="", x="", title="")
#Create a box plot of the amount of protein consumed in a day (DR1TPROT)

ggplot(NHANESdata, aes(x="", y=DR1TPROT))+ geom_boxplot()+ labs(y="total protein consumed daily in grams", x="", title="Box plot")
```


```{r}
#b. Create a box plot that comnpares a continuous variable across different levels of a nominal or ordinal variable: DR1TPROT by race:
#ggplot(data, aes(x=Independent variable, y=Contvar)) + geom_boxplot(fill="color") + labs(x="Independent variable", y="Contvar", title="")

ggplot(NHANESdata, aes(x=Race, y=DR1TPROT))+ geom_boxplot(fill="pink")+ labs(x="Race", y="DR1PROT", title="daily protein intake in grams")

```

